---
title: "10A: Processing and Visualising Flow Data"
subtitle: "In this exercise, we will explore the concept of spatial interaction, and learn how to build an OD (origin/destination) matrix."
# draft: true
date: "Sep 29, 2024"
date-modified: "last-modified"
author: Teng Kok Wai (Walter)
execute:
  echo: true
  eval: true
  freeze: true
  message: false
  warning: false
format:
  html:
    code-link: true
    toc: true
number-sections: true
number-offset: 1
editor: visual
---

## Exercise 10A Reference

[R for Geospatial Data Science and Analytics - 15  Processing and Visualising Flow Data](https://r4gdsa.netlify.app/chap15)

## Overview

In this exercise, we will explore the concept of spatial interaction, and learn how to build an OD (origin/destination) matrix.

Spatial interaction represent the flow of people, material, or information between locations in geographical space. It encompasses everything from freight shipments, energy flows, and the global trade in rare antiquities, to flight schedules, rush hour woes, and pedestrian foot traffic.

> An OD matrix, or spatial interaction matrix, represents each spatial interaction as a discrete origin/destination pair, where each pair corresponds to a cell in the matrix; the rows denote the locations (centroids) of origin, and the columns represent the locations (centroids) of destination.

## Learning Outcome

-   Import and extract OD data for a selected time interval.
-   Import and save geospatial data (bus stops and planning subzones) into sf tibble data frame objects.
-   Populate planning subzone codes into bus stops sf tibble data frames.
-   Construct desire lines geospatial data from the OD data.
-   Visualize passenger volume by origin and destination bus stops using the desire lines data.

## The Data

The following datasets will be used in this exercise:

| **Data Set** | **Description** | **Format** |
|------------------|-------------------------------------|------------------|
| Passenger Volume by Origin Destination Bus Stops | OD data set representing the volume of passengers traveling between bus stops. | CSV |
| BusStop | Geospatial data providing the locations of bus stops as of the last quarter of 2022. | ESRI Shapefile |
| MPSZ-2019 | Geospatial data providing the sub-zone boundary of the URA Master Plan 2019. | ESRI Shapefile |

## Installing and Launching the R Packages

The following R packages will be used in this exercise:

| **Package** | **Purpose** | **Use Case in Exercise** |
|------------------|---------------------------|---------------------------|
| **sf** | Handles vector-based geospatial data. | Importing, processing, and transforming geospatial data, such as bus stop locations and sub-zone boundaries. |
| **tidyverse** | A collection of packages for data science tasks such as data manipulation, visualization, and modeling. | Importing and wrangling OD and geospatial data, and visualizing analysis outputs. |
| **tmap** | Creates static and interactive thematic maps using cartographic quality elements. | Visualizing passenger flows and geographic clusters in a cartographic format. |
| **stplanr** | Provides functions for transport planning and modeling. | Creating geographic desire lines from OD data and solving transport-related problems. |
| **DT** | Provides an R interface to the JavaScript library DataTables for interactive table display. | Displaying data tables in an interactive format within the HTML output. |

To install and load these packages, use the following code:

```{r pacman}
pacman::p_load(tmap, sf, DT, stplanr, tidyverse, knitr)
```

## Preparing the Flow Data

### Importing the OD data

First, we import the *Passenger Volume by Origin Destination Bus Stops* dataset using `read_csv()` from the **readr** package.

```{r}
odbus <- read_csv("data/aspatial/origin_destination_bus_202408.csv")
glimpse(odbus)
```

*odbus* tibble data frame shows that the values in `ORIGIN_PT_CODE` and `DESTINATON_PT_CODE` are in character data type, we will convert themm into factor data type.

```{r}
odbus$ORIGIN_PT_CODE <- as.factor(odbus$ORIGIN_PT_CODE)
odbus$DESTINATION_PT_CODE <- as.factor(odbus$DESTINATION_PT_CODE)
glimpse(odbus)
```

### Extracting the Study Data

For the purpose of this exercise, we extract commuting flows on weekdays between 6 and 9 a.m. and sum the trips.

```{r}
odbus6_9 <- odbus %>%
  filter(DAY_TYPE == "WEEKDAY") %>%
  filter(TIME_PER_HOUR >= 6 & TIME_PER_HOUR <= 9) %>%
  group_by(ORIGIN_PT_CODE, DESTINATION_PT_CODE) %>%
  summarise(TRIPS = sum(TOTAL_TRIPS))
```

The table below shows the head content of `odbus6_9`:

```{r}
datatable(head(odbus6_9, 10))
```

### Saving and loading the data

We save the filtered data for future use in RDS format.

```{r}
#| eval: false

write_rds(odbus6_9, "data/rds/odbus6_9.rds")
odbus6_9 <- read_rds("data/rds/odbus6_9.rds")
```

## Working with Geospatial Data

For this exercise, two geospatial datasets will be used:

-   **BusStop**: Contains the locations of bus stops as of Q4 2022.
-   **MPSZ-2019**: Provides the sub-zone boundaries from the URA Master Plan 2019.

Both datasets are in ESRI shapefile format.

### Importing Geospatial Data

The code below imports the geospatial data:

```{r}
busstop <- st_read(dsn = "data/geospatial", layer = "BusStop") %>%
  st_transform(crs = 3414)
```

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MPSZ-2019") %>%
  st_transform(crs = 3414)
```

```{r}
mpsz <- write_rds(mpsz, "data/rds/mpsz.rds")
```

## Geospatial Data Wrangling

### Combining BusStop and MPSZ

The code below joins the planning subzone codes from `mpsz` to the bus stops in `busstop`:

```{r}
busstop_mpsz <- st_intersection(busstop, mpsz) %>%
  select(BUS_STOP_N, SUBZONE_C) %>%
  st_drop_geometry()
```

```{r}
datatable(head(busstop_mpsz, 10))
```

```{r}
nrow(busstop)
nrow(busstop_mpsz)
```

::: callout-note
-   `st_intersection()` performs a point-and-polygon overlay.
-   `select()` keeps only `BUS_STOP_N` and `SUBZONE_C` fields.
-   Five bus stops outside Singapore are excluded.
:::

Save the result as an RDS file:

```{r}
#| eval: false

write_rds(busstop_mpsz, "data/rds/busstop_mpsz.rds")
```

Next, append the subzone codes to the `odbus6_9` dataset:

```{r}
od_data <- left_join(odbus6_9, busstop_mpsz, by = c("ORIGIN_PT_CODE" = "BUS_STOP_N")) %>%
  rename(ORIGIN_BS = ORIGIN_PT_CODE, ORIGIN_SZ = SUBZONE_C, DESTIN_BS = DESTINATION_PT_CODE)
```

Check for duplicate records:

```{r}
duplicate <- od_data %>%
  group_by_all() %>%
  filter(n() > 1) %>%
  ungroup()

duplicate
```

Since duplicates exist, we will remove them:

```{r}
nrow(od_data)
od_data <- unique(od_data)
nrow(od_data)
```

Now, append the destination subzone codes:

```{r}
od_data <- left_join(od_data, busstop_mpsz, by = c("DESTIN_BS" = "BUS_STOP_N")) %>%
  rename(DESTIN_SZ = SUBZONE_C) %>%
  drop_na() %>%
  group_by(ORIGIN_SZ, DESTIN_SZ) %>%
  summarise(MORNING_PEAK = sum(TRIPS))

od_data_fii <- od_data
```

Finally, save the cleaned data:

```{r}
#| eval: false

write_rds(od_data_fii, "data/rds/od_data_fii.rds")
od_data_fii <- read_rds("data/rds/od_data_fii.rds")
```

## Visualising Spatial Interaction

In this section, we will prepare a desire line by using **stplanr** package.

### Removing Intra-Zonal Flows

We will exclude flows within the same zone to focus on inter-zonal flows. The code below removes these intra-zonal flows:

```{r}
#| eval: false
od_data_fij <- od_data[od_data$ORIGIN_SZ != od_data$DESTIN_SZ,]
```

::: callout-tip
The comma in the code `od_data1 <- od_data[od_data$ORIGIN_SZ != od_data$DESTIN_SZ,]` is crucial because it specifies that you are subsetting the rows of the data frame based on a condition, while keeping all the columns.
:::

Save the result for future use:

```{r}
#| eval: false

write_rds(od_data_fij, "data/rds/od_data_fij.rds")
```

```{r}
od_data_fij <- read_rds("data/rds/od_data_fij.rds")
```

### Creating Desire Lines

::: callout-note
Desire lines are used to illustrate on a map the flows of people or goods from point to point based on the values from a matrix.
:::

Next, we use `od2line()` from the **stplanr** package to generate desire lines:

```{r}
flowLine <- od2line(flow = od_data_fij, 
                    zones = mpsz,
                    zone_code = "SUBZONE_C")

flowLine
```

Save the generated desire lines:

```{r}
#| eval: false

write_rds(flowLine, "data/rds/flowLine.rds")

flowLine <- read_rds("data/rds/flowLine.rds")
```

### Filtering High-Volume Flows

To simplify the visual output and focus on significant flows, we filter for high-volumes flows.

```{r}
summary(flowLine$MORNING_PEAK)
```

For example, we can visualize flow greater than or equal to 2000 as shown below.

```{r viz_high_flow}
tm_shape(mpsz) +
  tm_polygons() +
flowLine %>%  
  # filter for 2000
  filter(MORNING_PEAK >= 2000) %>%  
tm_shape() +
  tm_lines(lwd = "MORNING_PEAK",
           style = "quantile",
           scale = c(0.1, 1, 3, 5, 7, 10),
           n = 6)
#           alpha = 0.3)
```
