---
title: "In-Class Exercise 1"
date: "Aug 26, 2024"
date-modified: "last-modified"
author: Teng Kok Wai (Walter)
execute:
  echo: true
  eval: true
  freeze: true
  message: false   
  warning: false
format:
  html:
    toc: true
number-sections: true
number-offset: 1    
editor: visual
---

## Import R Packages

```{r}
pacman::p_load(tidyverse, sf, tmap, ggstatsplot)
```

## Working with Master Plan Planning Sub-zone Data

To import the shapefile version of the Sub-zone data:

```{r}
mpsz14_shp <- st_read(dsn = "data", 
                  layer = "MP14_SUBZONE_WEB_PL")
```

Notice that `mpsz14_shp` is of class `sf` and inherits from `data.frame`. This means it combines spatial features with tabular data structure.

```{r}
class(mpsz14_shp)
```

To import the kml version of the Sub-zone data:

```{r}
#| eval: false
mpsz14_kml <- st_read("data/MasterPlan2014SubzoneBoundaryWebKML.kml")
```

Note that this is a corrupted file.

To render a workable kml, we can convert the shp object and save it as a kml file.

```{r}
#| output: false
st_write(mpsz14_shp,
         "data/MasterPlan2014SubzoneBoundaryWebKML.kml",
         delete_dsn=TRUE)
```

Note that the `delete_dsn=TRUE` argument ensures that any existing file with the same name is deleted before saving the new KML file, preventing potential conflicts or errors from overwriting.

## To import MPSZ-2019

```{r}
mpsz19_shp <- st_read(dsn = "data", 
                  layer = "MPSZ-2019") %>%
  st_transform(crs=3414)
```


```{r}
mpsz19_kml <- st_read("data/MasterPlan2019RegionBoundaryNoSeaKML.kml")
```
